Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-26 paper.tex
writing

The singular value expansion is useful for two reasons.  Firstly, the discrete version, the singular value decomposition, is the most stable way to solve badly-posed least squares problems.  Secondly, the sizes of the singular values, and the components of $H|ψ〉$ along the singular directions, determine whether the problem can be solved at all.

@@

@@

>>>>

@@

%% Tychonov conditioning

There is a family of methods for inverting operators with small singular values, known as regularisation methods.  A very useful idea is due to Tychonov.  The problem
$$\Gx f=|ψ〉$$
can be regarded as a least squares problem, where $|ψ〉$ is chosen to minimise the residual $‖\Gx f-|ψ〉‖²$.  The exact solution has residual zero, of course.  The problem is that it is swamped by components with small singular values of $\Gx$.  Suppose we modify this problem, and seek $|ψ〉$ to minimise $‖\Gx f-|ψ〉‖²+λ‖f‖²$, where $λ$ is a positive number, known as a Lagrange parameter for reasons that will become clear shortly.  There are two things to note here.  Firstly, this is still a least squares problem: it is equivalent to minimising 
$$\left‖ \pmatrix{\Gx \cr λ} f-\pmatrix{|ψ〉\cr 0}\right‖².$$
Therefore, it can be solved stably by the usual least squares methods.  Secondly, suppose $f_λ$ is the solution for a given value of $λ$.  If another function, $g$, had the same norm but a smaller residual, then $f$ would not minimise the Tychonov expression.  The Tychonov method is equivalent to constraining the norm of $f$, and seeking the solution with smallest residual given the constraint.  Or, equivalently, constraining the residual and seeking the solution with smallest norm.  This equivalence to a constrained optimisation problem is why $λ$ is a Lagrange parameter.

Solving the Tychonov problem for various values of $λ$ generates a family of solutions.  These can be plotted on a graph of residual against solution norm, where they mark the edge of the space of possible solutions.  Above and to the right of the curve, a solution can be found with the given norm and residual: below and to the left, it can't.

\topinsert\XeTeXpicfile lcurve-1.pdf \endinsert

An example for a typical application is shown below.  The context is that some observations $b$ are predicted by a physical state $x$, according to the linear law $Ax=b$.  The real state is $x₀$, and it has generated observations $b=Ax₀+e$, where $e$ consists of measurement error.  The problem $Ax=b$ is solved for $x$, with various Lagrange parameters.  At the top left, the norm of $x$ is constrained to be zero, so the solution is $x=0$, and the residual is the norm of $b$.  As the constraint on the norm is relaxed, the solution becomes closer to $x₀$, and the residual decreases.  At the elbow of the curve, the norm of the solution is constrained to be close to the norm of the real state $x₀$, the solution is close to $x₀$, and the residual is the norm of the measurement error.  Relaxing the constraint further allows solutions $x$ that fit the measurements better than the real state $x₀$ does.  Typically, they won't fit much better, so the residual doesn't decrease much below $‖e‖$.  However, the solution now includes components belonging to small singular values, and the norm of the solution will increase as far as it is allowed to.

@@

%% Tychonov regularisation
